---
title: "Predictors of Early Childhood Development in Nigeria Using Random Forest Models"
author: "Joshua W. Edefo"
date: "`r Sys.Date()`"
license: GPL-3.0
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  github_document: default
---
#Introduction
This study applies machine-learning methods to identify predictors
#of early childhood development outcomes in Nigeria using nationally representative survey data.


# ------------------------------------------------------------ 
# License: GPL-3.0

Copyright (C) 2025 Joshua W. Edefo

This program is free code: you can redistribute it and/or modify it under the terms of the GNU General Public License either version 3 of the License or any later version.

This code is distributed in the hope that it will be useful, but WITHOUT ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU General Public License for more details.

You should have received a copy of the GNU General Public License along with this program. If not, see <https://www.gnu.org/licenses/>.
# ------------------------------------------------------------ 

```{r a, message=FALSE}
library(haven)        # read_dta()
library(data.table)  # fast data manipulation, :=, .SD
library(fst)          # write_fst(), read_fst()
library(caret)        # train(), createDataPartition(), confusionMatrix()
library(ranger)       # random forest engine
library(ggplot2)      # plots
library(pROC)         # ROC / AUC
library(rcompanion)  # cramerV()
library(DescTools)   # cramerV()
library(usethis)    # project-setup package

set.seed(123)

# External data location
data_path <- "C:/Users/edefoj/Documents/Per/MICS/Nigeria _2021"
dta_file  <- file.path(data_path, "children_pro.dta")
fst_file  <- file.path(data_path, "children_pro.fst")
# Safety check
if (!file.exists(dta_file)) {
  stop("Stata file not found at: ", dta_file)
}

# Read the Stata file
data <- read_dta(dta_file)
setDT(data)

# Inspect data
head(data)
class(data)

# Select specified columns
data <- data[, .(agegrp1_cat, HL4, agegrp, HH6, zone, windex5MICS, cdisability, melevel_comb, EC1, ethnicity,UCD2G, UCD2K, UCF18, CA1)]


# Inspect structure
str(data)
nrow(data)
names(data)
head(data)
class(data)

# Modeling libraries
library(caret)
library(ranger)

# Target distribution
unique(data$agegrp1_cat)
prop.table(table(data$agegrp1_cat))

# Missing values
any(is.na(data))

############################################
# Data cleaning for ML (BASE R — SAFE)
############################################
# Ordinal predictors
ordinal_vars <- c("agegrp", "windex5MICS", "melevel_comb", "EC1")

#Convert numeric predictors to factors (excluding target)
for (v in ordinal_vars) {
data[[v]] <- factor(data[[v]], ordered = TRUE)
}

num_vars <- names(data)[sapply(data, is.numeric)]
num_vars <- setdiff(num_vars, "agegrp1_cat")

for (v in num_vars) {
data[[v]] <- factor(data[[v]])
}

#Target variable
data$agegrp1_cat <- factor(data$agegrp1_cat)

############################################
# Class balancing
############################################

tbl <- table(data$agegrp1_cat)
w <- 1 / tbl
w <- w / min(w)

############################################
# Feature exploration
############################################

library(ggplot2)
ggplot(data, aes(x = ethnicity)) + geom_bar()
ggplot(data, aes(x = agegrp)) + geom_bar()

############################################
# Association checks
############################################

library(rcompanion)

cat_cols <- names(Filter(is.factor, data))

cramer_matrix <- matrix(
  NA, length(cat_cols), length(cat_cols),
  dimnames = list(cat_cols, cat_cols)
)

for (i in cat_cols) {
  for (j in cat_cols) {
    cramer_matrix[i, j] <- cramerV(
      table(data[[i]], data[[j]]),
      bias.correct = TRUE
    )
  }
}

############################################
# Leakage check
############################################

target_var <- "agegrp1_cat"
predictors_list <- setdiff(names(data), target_var)

library(DescTools)
cramerV_target <- sapply(
  data[, ..predictors_list],
  function(x) cramerV(x, data[[target_var]])
)

sort(cramerV_target, decreasing = TRUE)

############################################
# Train/test split
############################################

set.seed(123)

test_index <- createDataPartition(data$agegrp1_cat, p = 0.2, list = FALSE)
test_data  <- data[test_index]
train_data <- data[-test_index]

train_weights <- w[train_data$agegrp1_cat]

############################################
# Fix factor levels
############################################

train_data$agegrp1_cat <- factor(
  train_data$agegrp1_cat,
  levels = c("0", "1"),
  labels = c("class0", "class1")
)

test_data$agegrp1_cat <- factor(
  test_data$agegrp1_cat,
  levels = c("0", "1"),
  labels = c("class0", "class1")
)

############################################
# Cross-validation
############################################

tc <- trainControl(
  method = "cv",
  number = 10,
  savePredictions = "final",
  classProbs = TRUE,
  allowParallel = TRUE
)

############################################
# Default ranger model
############################################

default_model <- train(
  agegrp1_cat ~ .,
  data = train_data,
  method = "ranger",
  trControl = tc,
  tuneGrid = data.frame(
    mtry = floor(sqrt(ncol(train_data) - 1)),
    splitrule = "gini",
    min.node.size = 1
  ),
  weights = train_weights,
  importance = "impurity"
)

############################################
# Tuned ranger model
############################################

tuned_grid <- expand.grid(
  mtry = c(3, 5, 7),
  splitrule = c("gini", "extratrees"),
  min.node.size = c(1, 5, 10)
)

tuned_model <- train(
  agegrp1_cat ~ .,
  data = train_data,
  method = "ranger",
  trControl = tc,
  tuneGrid = tuned_grid,
  weights = train_weights,
  importance = "impurity"
)

############################################
# Model selection
############################################

best_model <- if (max(tuned_model$results$Accuracy) >
                  max(default_model$results$Accuracy)) {
  tuned_model
} else {
  default_model
}

############################################
# Test performance
############################################

test_pred <- predict(best_model, test_data)
mean(test_pred == test_data$agegrp1_cat)

confusionMatrix(
  test_pred,
  test_data$agegrp1_cat,
  positive = "class1"
)

```

 Model perfoemance Evaluatioons and plots

```{r c}

############################################
# ROC / AUC
############################################

library(pROC)
test_prob <- predict(best_model, test_data, type = "prob")[, "class1"]
roc_obj <- roc(test_data$agegrp1_cat, test_prob)
auc(roc_obj)
plot(roc_obj)

############################################
# Feature importance
############################################
# Feature importance
imp <- best_model$finalModel$variable.importance

imp_df <- data.frame(
  Feature = names(imp),
  Importance = imp
)

imp_df <- imp_df[order(imp_df$Importance, decreasing = TRUE), ]

imp_df2 <- aggregate(
  Importance ~ gsub("\\..*|\\^.*", "", Feature),
  imp_df,
  sum
)

names(imp_df2)[1] <- "Feature"

library(ggplot2)

ggplot(imp_df2, aes(x = reorder(Feature, Importance), y = Importance)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Variable Importance (Aggregated)",
    x = "Predictor",
    y = "Importance"
  )

```
## Session information

```{r session-info, message=FALSE, warning=FALSE}

sessionInfo()
```
